Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
Check in of preliminary 2-term intro to splitting.  Now working on 1M

\input{figures/splitting-overview}

@@

%Although the type system of \lang\ ensures that we can perform all first-stage
%computations before any second-stage ones, actually evaluating terms in this
%fashion is rather involved, because stages are syntactically interleaved
%within terms.

In this section, we present a {\em splitting algorithm} that
statically converts a term in \lang\ into a pair of monostaged terms
that individually correspond to the first and second stage
computations of the original \lang\ term. Due to important differences
in structure, we describe both the invariants and algorithm for
splitting \bbonem\ terms separately from that of \bbtwo\ terms.  By
construction, \bbonep\ terms are monostage and do not require
splitting.

\textbf{\bbtwo\ term splitting.}  As suggested informally in
\ref{sec:example}, splitting a \lang\ program (a \bbtwo\ term) should
yield two monostage terms where evaluating the second term
(corresponding to second stage computations) on the output of the
first term (first stage computations), yields the same result as
direct evaluation of the original \lang\ program.  More precisely, for
any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits (via the new relation~$\splittwosym$) into a pair of terms
$\pipeS p l r$ where $\reduce p b$, and $[b/l]r$ is the same as $q$ up
to an equivalence. The term $p$ generated by the split (called the
{\em precomputation}, due to its role producing the inputs for
subsequent second stage computation) is a monostage term representing
all the first-stage computations in $e$. The term $l.r$ (called the
{\em resumer}) is a monostage term representing all the second-stage
computations in $e$.  We refer to the result of the precomputation
($b$) the {\em boundary value} of the split since it represents the
data communicated between the first and second stage computations.
The process of splitting \bbtwo\ terms (and the relationship of these
operations to the outputs of direct \lang\ evaluation) is provided at
right in \ref{fig:splittingSummary}.
% and described in detail in \ref{sec:split-two}.

@@

\textbf{\bbonem\ term splitting.} Defining the requirements for
splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
(which reduce to monostage residuals that can be directly compared to
resumers), \bbonem\ terms reduce to residual tables and partial values
that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
%For example, the \bbonem\ term:
%\begin{lstlisting}
%(next {1+2}, grnd{3+4})
%\end{lstlisting}
%reduces (via $\redonesym$) to the residual table and partial value:
%\begin{lstlisting}
%[yhat |-> 1+2] (next {yhat}, grnd{7})
%\end{lstlisting}

@@

We define the requirements for splitting \bbonem\ terms using the
results of masking. Specifically, for any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
equivalence.  The term $c$ generated by splitting (called the {\em
  combined term}) is a monostage term encoding all the first-stage
computations in $e$, and $l.r$ (again called the {\em resumer}) is a
monostage term encoding all the second-stage computations in $e$.  The
process of splitting \bbonem\ terms is summarized at left in
\ref{fig:splittingSummary}.
% and described in detail in \ref{sec:split-one}.

%$\sepredonesym$ is defined by a single rule that yields masked values:
%\[
%\infer{\sepredone {\pipeM c l r} {\mval i {[b/l]r}}} {\reduce c {(i,b)}}
%\]

>>>>

%\input{figures/splitting-overview}

@@

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms that correspond to the first stage and second stage
computations in the original term. The idea is that evaluating the
first monostaged term in a first pass, and then evaluating the second
monostaged term on the results of the first pass, yields the same
result as fully evaluating the original \lang\ program.  Complexity
arises because \lang\ evaluation of \bbonem\ terms may also generate
results in the first stage. (For example, a \bbonem\ term of type
$\curr \rmint$ results in an \rmint\ in the first stage.) Handling the
asymmetry between the results of evaluating \bbonem\ terms (which may
produce results at both the first and second stage) and \bbtwo\ terms
(which only produce second stage results) is fundamental to the
structure of the splitting algorithm presented in this section.  By
construction, \bbonep\ terms are monostage and do not require
splitting.

\textbf{\bbtwo\ term splitting.}  For any $\coltwo e A$ which reduces
to a residual $q$ (via $\diatwo e q$) and then a final value $v$ (via
$\reduce q v$), splitting $e$ (via the new relation~$\splittwosym$)
yields a pair of monostaged terms $p$ (containing first stage
computations in $e$) and $l.r$ (containing second stage computations
in $e$). Executing $p$ in a first pass produces $b$ (via $\reduce p
b$). Then, in a second pass, $\reduce {[b/l]r} {v'} = v$.  Since
execution of the first pass serves to generate input for the second
pass, we say $p$ is a {\em precomputation} that produces a {\em
  boundary value} ($b$) that is communicated to the {\em resumer}
($l.r$).

The process of splitting \bbtwo\ terms is given by via the new
relation~$\splittwosym$.

% (and the relationship of these
% operations to the outputs of direct \lang\ evaluation) is provided at
% right in \ref{fig:splittingSummary}.

@@

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
equivalence.  The term $c$ generated by splitting (called the {\em
  combined term}) is a monostage term encoding all the first-stage
computations in $e$, and $l.r$ (again called the {\em resumer}) is a
monostage term encoding all the second-stage computations in $e$.  The
process of splitting \bbonem\ terms is summarized at left in
\ref{fig:splittingSummary}.

%Defining the requirements for
%splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
%(which reduce to monostage residuals that can be directly compared to
%resumers), \bbonem\ terms reduce to residual tables and partial values
%that represent computations at both stages.

@@

% We define the requirements for splitting \bbonem\ terms using the
% results of masking.
